---
title: "ggplot2: Adding more information based on variables in your data"
format: html
editor: visual
---

# Mappings

In `ggplot2`, the mapping specifies how variables (columns) are mapped onto aesthetic attributes. The mapping between variables and aesthetics is accomplished with the `aes` function. Example:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point()
```

## Adding aesthetics

Besides x and y, you may want to map variables to other aesthetic attributes: color, fill, shape, line type, size, line width, label, etc. This is useful, for example, to show how a relationship between two numeric variables varies by a categorical variable.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight, color = gender)) +
  geom_point()
```

You can even have two categorical variables at the same time, although it can quickly become too much for one plot.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight, color = gender, shape = physically_active)) +
  geom_point()
```

### TRY IT

Create a scatter plot (`geom_point`) of `cholesterol` (`y`) against `weight` (`x`).

```{r}

```

Map `gender` to the `color` aesthetic.

```{r}

```
