---
title: "ggplot2: Plotting subsets of the data"
format: html
editor: visual
---

# Facets

Sometimes you want to plot different subsets of the data using the same visualization settings. Facets help you do this with functions that start with `facet_`.

[`facet_grid`](https://ggplot2.tidyverse.org/reference/facet_grid.html) is the most common function to add facets to a plot. It forms a matrix of panels defined by row and column faceting variables.

Here we're using `rows` to specify that we want each value of `physically_active` to be a different row (as opposed to `cols`). We need to use `vars` to specify the variable to use for the facets. Think of `vars` as `aes` -- it's the way to tell `ggplot2` that you want to use columns from the data set specified in `ggplot`.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(bins = 20) +
  facet_grid(rows = vars(physically_active))
```

## TRY IT

Facet the plot below by `gender`, with each `gender` being a different column.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(bins = 20)
```
